---
title: "Modules"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

<button class="accordion">Before Starting the Program</button>
<div class="panel"> 

| Topic                              | Links |
|------------------------------------|-------|
| Download and Install R | [Link](https://mirror.las.iastate.edu/CRAN/)|
| Download and Install RStudio| [Link](https://www.rstudio.com/products/rstudio/download/)|
|Download packages |<a href="R-code/code-00-setup.R" download>File</a>|
|Alzheimer's Disease | [Link](https://www.youtube.com/watch?v=LrcL3vJjN2g) |

</div>




<button class="accordion">Day 1 &#9679; Data Science -- Basics and Principles  </button>
<div class="panel">

| Time          | Topic                              | Links |
|---------------|------------------------------------|-------|
| 9:00 - 9:15 | Welcome                            |[Slides](https://HenryYang03.github.io/ics_summer_files/Welcome.html)|
| 9:15 - 9:45 | Introduction to data science    |[Slides](https://uci-data-academy.github.io/slides/01b-intro-data-science.html)|
| 10:00 - 10:45 | Describing data    |[Slides](https://uci-data-academy.github.io/slides/01c-describe-data.html)   |
| 11:00 - 12:00 | Introduction to the capstone project |                [Slides](https://uci-data-academy.github.io/slides/2023_Intro_to_AD_Online.pdf)    |
| 1:30 - 2:30 | Introduction to R |[Slides](https://uci-data-academy.github.io/slides/Lab-01a-intro-r-schetzsle.html) |
| 2:45 - 3:30 | R packages |[Slides](https://uci-data-academy.github.io/slides/Lab-01b-R-packages.html) |
</div>

	
	
```{r}
htmltools::tagList(
  htmltools::tags$script(src = "https://d3js.org/d3.v4.min.js"),
  htmltools::tags$script(src = "https://cdnjs.cloudflare.com/ajax/libs/dimple/2.3.0/dimple.latest.min.js")
)
```

```{js, class.source = "jsvis1"}
var acc = document.getElementsByClassName("accordion");
var i;

for (i = 0; i < acc.length; i++) {
  acc[i].addEventListener("click", function() {
    /* Toggle between adding and removing the "active" class,
    to highlight the button that controls the panel */
    this.classList.toggle("active");

    /* Toggle between hiding and showing the active panel */
    var panel = this.nextElementSibling;
    if (panel.style.display === "block") {
      panel.style.display = "none";
    } else {
      panel.style.display = "block";
    }
  });
}
```
