---
title: "Day 1: Describe Data"
author: "Zhaoxia Yu, Department of Statistics, University of California, Irvine"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: "sky"
    slideNumber: true
    transition: "fade"
    progress: true
    controls: true
---

# Introduction

## Describe Data

- Describing data is foundational step in the data science life cycle
- It involves summarizing and visualizing the main characteristics of a data set.

## Describe Data: Key Roles
  - **Understanding Data**: Gain insights into the data's structure, distribution, and underlying patterns.
  - **Informing Analysis**: Identify important variables, potential outliers, and data anomalies that need attention.
  - **Guiding Data Cleaning**: Detect missing values, inconsistencies, and errors that must be addressed before further analysis.

## Describe Data: Descriptive Techniques
  - **Summary Statistics**: Measures of central tendency (mean, median) and dispersion (range, variance).
  - **Data Visualization**: Graphical methods such as histograms, bar charts, box plots, and scatter plots.
  - **Exploratory Data Analysis (EDA)**: Techniques to explore data and identify patterns, relationships, and anomalies.


## Describe Data:  Impact
  - **Data Collection**: Ensures the quality and relevance of collected data.
  - **Data Cleaning**: Helps identify issues in the data.
  - **Data Analysis**: Informs the choice of analysis methods and models.
  - **Model Building**: Provides insights into feature selection and engineering.
  - **Decision Making**: Supports data-driven decisions by providing a clear understanding of the data.


## Some Terminologies

- When studying a population, we measure various characteristics, known as __variables__.

- The goal of many scientific studies is to understand the __variation__ of specific characteristics (e.g., BMI, disease status) within the population of interest.

- Often, studies focus on exploring potential __relationships__ between different variables.

## Example: Cognitive Status and Brain Volume
- A subset of data from the **Uniform Data Set** of National Alzheimer’s Coordinating Center (NACC)
- Measurements of brain volume
- Cognitive status: normal cognition (0), mild cognitive impairment due to AD (1), or dementia due to AD (2)
- Other important information, such as age and gender
- We will be using this data set to illustrate how to use different statistical methods to address questions in data science 

## A Subset of the Data 
```{r}
#| warning: false
#| code-fold: true
#| echo: true

library(tidyverse)

#old-fashioned way to handle data
#step 1: read the file
alzheimer_data <- read.csv('data/alzheimer_data.csv')
#dim(alzheimer_data) #check the dimension 

#step 2: take a subset of columns of interest
alzheimer_data = alzheimer_data[, c('id', 'diagnosis', 'age', 'educ', 'female', 'height', 'weight', 'lhippo', 'rhippo')]
#dim(alzheimer_data) #check the dimension

#step 3: change diagnosis and gender to factor variables
alzheimer_data[,c("diagnosis")]=as.factor(alzheimer_data[,c("diagnosis")])
alzheimer_data[,c("female")]=as.factor(alzheimer_data[,c("female")])

#use cool functions from the tidyverse package
alzheimer_data <- read.csv('data/alzheimer_data.csv') %>% 
  select(id, diagnosis, age, educ, female, height, weight, lhippo, rhippo) %>% 
  mutate(diagnosis = as.factor(diagnosis), female = as.factor(female))
head(alzheimer_data)
```


# Variable Types

## Variable Types - Response vs Explanatory Variable

- Variables that are the primary focus of our study are called __response__ (or target) variables. They are also called __dependent__ variables. 

- Variables that help explain or predict the variation in the response variable are known as __explanatory/independent__ variables or __predictors__, depending on their role.


- Statistical analysis typically starts with a scientific question, often framed as a __hypothesis testing__, __estimation__, or __prediction__ problem.

## Variable Type: Numerical

- A **Numerical Variable** takes numerical values that hold their usual **quantitative** meaning.

- Example: `age`, `height`, `weight`, `lhippo`, and `rhippo` take numerical values that hold their usual quantitative meaning.
- These variables allow for meaningful analyses, such as calculating averages, and determining maximum or minimum values.

- Not all numerical representations indicate a numeric quantity. For example, `diagnosis` may use numbers.


## Variable Type: Categorical
- A **categorical variable**, also known as a discrete or qualitative variable, consists of a limited number of categories. 

- Example: `diagnosis` and `female`. In this data set, `female' is a binary variable.

- Sometimes numerical codes are assigned to categorical variables, but these numbers do not carry their typical quantitative meaning.



## Describe a Variable

- We begin by exploring data using techniques that examine one variable at a time.

- Our goal is to gain a high-level understanding of the data, learn about the potential values for each characteristic, and discover how each characteristic varies among individuals in our sample.

- Essentially, we aim to understand the __distribution__ of the variables.

# Categorical Variables

## Describe a Categorical Variable 

- **Frequency Distribution**: the count of observations in each category. The frequency for category $c$ by is often denoted by $n_{c}$.

- **Relative Frequencies**: the relative frequency of category $c$ is defined as 

$$
\begin{equation*}
p_{c} = \frac{n_{c}}{n}
\end{equation*}
$$

- For a categorical variable, the __mode__ is the most common value, i.e.,
the value with the highest frequency.


## Describe a Categorical Variable 

- **Bar Charts**: Visual representation showing frequencies or proportions of each category.

- Example

```{r echo=FALSE, out.width='45%', fig.align='center'}
#| echo: true
#| code-fold: true
ggplot(alzheimer_data, aes(diagnosis)) +
  geom_bar()
```

## A Nice Bar Plot
```{r echo=FALSE, out.width='45%', fig.align='center'}
#| echo: true
#| code-fold: true
ggplot(alzheimer_data, aes(diagnosis, fill=diagnosis)) +
  geom_bar()+
  scale_fill_discrete(labels=c('normal', 'impaired', 'AD'))
```

## Problem Solving
- How can you display relative frequencies (i.e., proportions), rather than counts?


# Numerical Variables

## Understanding Numerical Variables

- When analyzing numerical variables, our focus centers on two essential aspects of their distribution: __location__ and __spread__.

- __Location__: the location of a distribution refers to the _central tendency_ of values, that is, the point around which most values are gathered. 

- __Spread__: refers to the _dispersion_ of possible values, that is, how
scattered the values are around the location.

## Example

```{r, out.width="100%", fig.align="center"}
#| echo: true
#| code-fold: true
# Generating example data
set.seed(321)
data1 <- rnorm(100, mean = 0, sd = 1)   # Distribution 1: Mean 0, SD 1
data2 <- rnorm(100, mean = 0, sd = 2)   # Distribution 2: Mean 0, SD 2
data3 <- rnorm(100, mean = -2, sd = 1)  # Distribution 3: Mean -2, SD 1

# Creating a data frame for plotting
df <- data.frame(
  Group = rep(c("Group 1", "Group 2", "Group 3"), each = 100),
  Value = c(data1, data2, data3)
)

# Plotting one-dimensional scatter plots (strip plots)
library(ggplot2)
ggplot(df, aes(x = Group, y = Value, color = Group)) +
  geom_point()+
  labs(title = "Distributions with Different Locations and Spreads",
       x = "Group", y = "Value") +
  theme_minimal()
```

## Histogram

- A histogram resembles a bar graph, where the variable's values are grouped (binned) into a set number of intervals (bins).

- Each bar's height in the histogram indicates the frequency (count) of observations falling within that interval.

## Plotting histograms
```{r, out.width="100%", fig.align="center"}
#| echo: true
#| code-fold: true
histogram_plot <- ggplot(df, aes(x = Value, fill = Group)) +
  geom_histogram(binwidth = 0.5, position = "identity", alpha = 0.7) +
  labs(title = "Histograms of Distributions",
       x = "Value", y = "Frequency") +
  theme_minimal()

histogram_plot
```

## Histograms

```{r, out.width="100%", fig.align="center"}
#| echo: true
#| code-fold: true
library(gridExtra)

# Histograms
histogram_plot1 <- ggplot(df[df$Group == "Group 1", ], aes(x = Value)) +
  geom_histogram(binwidth = 0.5, fill = "skyblue", alpha = 0.7) +
  labs(title = "Histogram: Group 1",
       x = "Value", y = "Frequency") +
  theme_minimal()

histogram_plot2 <- ggplot(df[df$Group == "Group 2", ], aes(x = Value)) +
  geom_histogram(binwidth = 0.5, fill = "lightgreen", alpha = 0.7) +
  labs(title = "Histogram: Group 2",
       x = "Value", y = "Frequency") +
  theme_minimal()

histogram_plot3 <- ggplot(df[df$Group == "Group 3", ], aes(x = Value)) +
  geom_histogram(binwidth = 0.5, fill = "salmon", alpha = 0.7) +
  labs(title = "Histogram: Group 3",
       x = "Value", y = "Frequency") +
  theme_minimal()

# Combine plots into subplots
grid.arrange(histogram_plot1, histogram_plot2, histogram_plot3,
             ncol = 3, nrow = 1, top = "Scatter Plots and Histograms")
```

## Examples of Histograms
```{r, out.width="100%", fig.align="center"}
#| echo: true
#| code-fold: true
par(mfrow=c(1,2))
hist(alzheimer_data$age, xlab="age")
hist(alzheimer_data$lhippo, xlab="left hippo")

#The above code is old-fashioned R code without using ggplot2. Try to use ggplot to produce nicer plots.
```

## Histogram Shapes
- Beyond the location and spread of a distribution, the shape of a histogram reveals how the observed values are distributed around the center.

- A histogram is considered symmetric if the densities are nearly identical for any two intervals that are equidistant from the center.

- Often, histograms may be stretched to the left (left-skewed) or to the right (right-skewed), indicating asymmetry in the distribution.


## Important Summary Statistics
- Consider a sample of size $n$: $x_{1}, x_2, \ldots,x_{n}$. Here $n$ is called sample size. 
- The __sample mean__ is simply the average of the observed values:

$$
\bar{x} = \frac{\sum_{i=1}^n x_{i}}{n} = \frac{x_1 + x_2 + \cdots + x_n}{n},
$$

- The __sample median__ is an alternative measure of location, which
is less sensitive to outliers. 

  - Odd $n$, the median is the number at the middle of the sorted observations.

  - Even $n$, the median is the average of the two middle numbers.
  
## Standard Deviation and Variance

- Two widely used summary statistics for quantifying dispersion are the __sample variance__ and __sample standard deviation__.

- These statistics rely on the __deviation__ of observed values from the mean, which serves as the center of the distribution.

- For each observation, the deviation from the mean is calculated as $x_{i} - \bar{x}$.
  

## Standard deviation and Variance

- The sample variance is a common measure of dispersion based on the squared deviations
$$
\begin{equation*}
s^2  =  \frac{\sum_{i=1}^n (x_{i} - \bar{x} )^2 }{n-1}.
\end{equation*}
$$

- The square root of the variance is called the sample {standard deviation}

$$
\begin{equation*}
s  =  \sqrt{\frac{\sum_{i=1}^{n}(x_{i} - \bar{x})^{2}}{n-1}},
\end{equation*}
$$

## Example
```{r}
#| echo: true
#| code-fold: true
mean(alzheimer_data$age) #calculate mean
var(alzheimer_data$age) #calculate sample variance
sd(alzheimer_data$age) #calculate sample standard deviation
```

## Quantiles

- The median is the value that is greater than or equal to at least half of the values and less than or equal to the other half. The median is also known as the 0.5 __quantile__.

- Similarly, the 0.25 quantile is the value that is greater than or equal to at least 25% of the values.

- The 0.75 quantile is the value that is greater than or equal to at least 75% of the values.

- In general, the $q$ quantile is the value that is greater than or equal to at least $100q$% of the values.

- Sometimes, the $q$ quantile is referred to as the $100q$th __percentile__.

## Quartiles

- We can divide the ordered values of a variable into four equal parts
using 0.25, 0.5, and 0.75 quantiles. 

- The corresponding points are denoted $Q_{1}$, $Q_{2}$, and $Q_{3}$, respectively. 


- We refer to these three points as __quartiles__, of which $Q_{1}$ is called the _first quartile_ or the _lower quartile_, $Q_{2}$ (i.e., median)
is called the _second_ quartile, and $Q_{3}$ is called the _third quartile_ or _upper quartile_.

-  The interval from $Q_{1}$ (0.25 quantile) to $Q_{3}$ (0.75 quantile) covers the middle 50% of the ordered data.

## Five-Number Summary and Boxplot

- The __minimum__ (min), which is the smallest value of the variable in our sample, is the 0 quantile.

- Conversely, the __maximum__ (max), which is the largest value of the variable in our sample, is the 1 quantile.

- Together with the quartiles $Q_{1}$, $Q_{2}$, and $Q_{3}$, the minimum and maximum values constitute the __five-number summary__.

- These values are typically presented in ascending order: min, first quartile, median, third quartile, max.

- This five-number summary provides the 0, 0.25, 0.50, 0.75, and 1 quantiles.


## Example
```{r}
#| echo: true
#| code-fold: true
quantile(alzheimer_data$age) #calculate mean
```



## Five-Number Summary and Boxplot

- The five-number summary can be used to derive two measures of dispersion: the __range__ and the __interquartile range__ (IQR).

- The range is the difference between the maximum observed value and the minimum observed value.

- The interquartile range (IQR) is the difference between the third quartile $Q_3$ and the first quartile $Q_1$, represented by the formula: \(\text{IQR} = Q_3 - Q_1\). This measures the spread of the middle 50% of the

## Five-Number Summary and Boxplot

- The dashed lines extending from the box are referred to as the __whiskers__.

- The right whisker extends to the largest observed value or $Q_{3} + 1.5 \times \text{IQR}$, whichever is reached first.

- The left whisker extends to the lowest value or $Q_{1} - 1.5 \times \text{IQR}$, whichever is reached first.

- Data points beyond the whiskers are displayed as circles and are considered potential outliers. 

## Example: left hippocampus volume
```{r, out.width="100%", fig.align="center"}
#| echo: true
#| code-fold: true
# Create a boxplot of age
ggplot(alzheimer_data, aes(x = "", y = lhippo)) +  # x = "" for a single boxplot
  geom_boxplot(fill = "skyblue", color = "blue") +  # Customizing boxplot appearance
  labs(title = "Boxplot of Left Hippo in Alzheimer's Dataset", x="", y = "lhippo") +  # Labels for title and y-axis
  theme_minimal()  # Applying a minimal theme (optional, adjust as needed)
```

## Provlem Solving challenge
- How to draw boxplots of left hippo volume by diagnosis groups

- How to draw boxplots of left hippo volume by by diagnosis and gender

- What is a violin plot and how to draw one?

## Example: with two outliers
```{r, out.width="100%", fig.align="center"}
#| echo: true
#| code-fold: true
#generate data
x=rnorm(100) #100 data points from normal distribution with mean 0 and sd=1
x=c(x, 9.9, 9.8)#add two outliers
boxplot(x)
```

