---
title: "Day 2: Simulate Data"
author: "Sara Tyo, Department of Statistics, University of California, Irvine"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: "sky"
    slideNumber: true
    transition: "fade"
    progress: true
    controls: true
---

## What is Data Simulation?

-   **Definition:** Data simulation is the process of generating synthetic data that mimics real-world data characteristics.

-   Purposes:

    -   Exploration: Understand data distributions and relationships.

    -   Analysis: Test hypotheses, evaluate models, and simulate scenarios.

    -   Training: Prepare datasets for training machine learning models.

## Example

```{r, out.width="100%", fig.align="center"}
#| echo: true
#| code-fold: true
# Generating example data
set.seed(321)
data1 <- rnorm(100, mean = 0, sd = 1)   # Distribution 1: Mean 0, SD 1
data2 <- rnorm(100, mean = 0, sd = 2)   # Distribution 2: Mean 0, SD 2
data3 <- rnorm(100, mean = -2, sd = 1)  # Distribution 3: Mean -2, SD 1

# Creating a data frame for plotting
df <- data.frame(
  Group = rep(c("Group 1", "Group 2", "Group 3"), each = 100),
  Value = c(data1, data2, data3)
)
```

## **Explanation:**

### `set.seed()`

#### 1. **Purpose:**

-   **Reproducibility:** When you generate random numbers in R using functions like `rnorm()` (for normal distribution), `runif()` (for uniform distribution), etc., the numbers are pseudo-random. This means they appear random but are actually generated using a deterministic algorithm that starts from a specific initial state (seed).

## 

-   `rnorm`: Function in R to generate random numbers from a normal distribution.

-   Parameters:

    -   `mean`: Mean of the distribution.

    -   `sd`: Standard deviation of the distribution.

-   Generates three sets (`data1`, `data2`, `data3`) of 100 random numbers each with different means and standard deviations.

## Histograms

-   Show histograms of `data1`, `data2`, and `data3`.

-   Discuss the impact of mean and standard deviation on the shape and spread of the distributions.

```{r, out.width="100%", fig.align="center"}
#| echo: true
#| code-fold: true
library(ggplot2)
ggplot(df, aes(x = Value, fill = Group, stat="count")) +  
  geom_histogram() +  
  labs(title = "Distributions of Values across Groups",
       x = "Group", y = "Count") +
  theme_minimal()
```

## Importance of Simulation in Data Analysis

-   **Advantages:**

    -   **Flexibility:** Easily generate data to fit specific scenarios.

    -   **Control:** Adjust parameters to simulate various conditions.

    -   **Testing:** Validate statistical methods and models.

-   **Applications:**

    -   Financial modeling, risk assessment, predictive analytics, etc.

## Practical Applications

-   **Example:** Simulating customer purchase behavior in an online store.

    -   **Variables:** Purchase frequency, average order value, time between purchases.

    -   **Purpose:** Predict sales, optimize marketing strategies.

-   Possible way of simulating

    ```{r, out.width="100%", fig.align="center"}
    #| echo: true
    #| code-fold: true
    library(ggplot2)
    library(dplyr)

    # Set seed for reproducibility
    set.seed(123)

    # Number of customers
    num_customers <- 1000

    # Simulate purchase frequency
    purchase_frequency <- round(runif(num_customers, min = 1, max = 20))
    # Simulate average order value
    average_order_value <- rnorm(num_customers, mean = 50, sd = 10)
    # Simulate time between purchases (in days)
    time_between_purchases <- round(rnorm(num_customers, mean = 30, sd = 10))

    # Create a data frame
    customer_data <- data.frame(
      customer_id = seq(1, num_customers),
      purchase_frequency = purchase_frequency,
      average_order_value = average_order_value,
      time_between_purchases = time_between_purchases
    )
    ```

## Conclusion

-   Summary:

    -   Data Simulation as a Powerful Analytical Tool Data simulation stands as a cornerstone in modern data science, providing robust capabilities for understanding and analyzing complex datasets.

    -   By generating synthetic data that mirrors real-world scenarios, businesses and researchers alike can uncover hidden patterns, test hypotheses, and make informed decisions.
