---
title: "Day 3: Hypothesis Testing - Code Only"
author: "Zhaoxia Yu, Department of Statistics, University of California, Irvine"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: "sky"
    slideNumber: true
    transition: "fade"
    progress: true
    controls: true

execute:
  echo: true
---

# Load libraries
```{r}
#| code-fold: true
#| 
library(tidyverse)
library(ggplot2)
```


## Read Data

```{r}
#| code-fold: true

alzheimer_data <- read.csv('data/alzheimer_data.csv') %>% 
  select(id, diagnosis, age, educ, female, height, weight, lhippo, rhippo) %>% 
  mutate(diagnosis = as.factor(diagnosis), female = as.factor(female))


alzheimer_healthy <- read.csv('data/alzheimer_data.csv') %>% 
  select(id, diagnosis, age, educ, female, height, weight, lhippo, rhippo) %>% 
  mutate(diagnosis = as.factor(diagnosis), female = as.factor(female)) %>% filter(diagnosis==0)
```



# Example 1: One Sample t-test

## Example 1: visualization

## Example 1: caluclate mean
```{r}
#| code-fold: true
mean(alzheimer_healthy$lhippo)
```

## Example 1:  Calculate the test statistics

```{r}
#| code-fold: true
n=length(alzheimer_healthy$lhippo)
(mean(alzheimer_healthy$lhippo)-3)/ (sd(alzheimer_healthy$lhippo)/sqrt(n))
```


## Example 1: calculate p-value
```{r}
#| code-fold: true
#probability of greater than 17.72495
print("probability of greater than 17.72495")
pt(17.72495, df=n-1, lower.tail=F)

#probability of less than -17.72495
print("probability of less than -17.72495")
pt(-17.72495, df=n-1, lower.tail=T)

#p-value
print("p-value")
pt(17.72495, df=n-1, lower.tail=F)+pt(-17.72495, df=n-1, lower.tail=T)
```

## Example 1: Use t.test

We don’t have to the calculation manually. We can run a one-sample t-test using `t.test`

```{r}
#| code-fold: true

#the default test is two-tailed/sided
t.test(alzheimer_healthy$lhippo, mu=3)  
```

# EXAMPLE 1 L: LEFT-TAILED

```{r}
#| code-fold: true
#use the CDF of t to compute p-value
pt(17.72495, df=n-1, lower.tail=T)

#use t-test
t.test(alzheimer_healthy$lhippo, mu=3, alternative = "less")  
```

# EXAMPLE 1 R: RIGHT-TAILED

```{r}
#| code-fold: true
#use the CDF of t to compute p-value
pt(17.72495, df=n-1, lower.tail=F)
#use t-test
t.test(alzheimer_healthy$lhippo, mu=3, alternative = "greater")  
```


# Example 2: one-sample proportion z-test

## Example 2: Visualization

## Example 2: testing
```{r}
#| code-fold: true
phat=mean(alzheimer_healthy$lhippo>3 )
phat

phat.se=sqrt(phat*(1-phat)/n)
phat.se

(phat-0.5)/phat.se

#p-value
pnorm((phat-0.5)/phat.se, lower.tail = F)
```


# EXAMPLE 3: Two-Sample t-test

## Example 3: Visualization

## Example 3: use t.test
```{r}
#| code-fold: true
t.test(alzheimer_healthy$lhippo[alzheimer_healthy$female==0], alzheimer_healthy$lhippo[alzheimer_healthy$female==1])

t.test(alzheimer_healthy$lhippo ~ alzheimer_healthy$female)
```
  
  
#  EXAMPLE 4: paired t-test

## EXAMPLE 4: Visaulization 

## Example 4: use t.test
```{r}
#| code-fold: true
t.test(alzheimer_data$rhippo, alzheimer_data$lhippo, paired = T)

#equivalently, 
t.test(alzheimer_data$rhippo- alzheimer_data$lhippo)
```

# Example 5: chi-squared test

## Example 5: visualization

## Example 5: use chisq.test
```{r}
#| code-fold: true

table(alzheimer_data$rhippo>alzheimer_data$lhippo,
      alzheimer_data$female)

chisq.test(table(alzheimer_data$rhippo>alzheimer_data$lhippo,
      alzheimer_data$female))
```


# Example 6: Correlation

## Example 6: Visualization
```{r}
#| code-fold: true
#| fig-align: center

alzheimer_healthy$hippo = (alzheimer_healthy$lhippo+alzheimer_healthy$rhippo)/2

plot(alzheimer_healthy$age, alzheimer_healthy$hippo)
```
Discussion: is there a linear trend? are there regions of concerns? 


## Example 6: cor.test
```{r}
#| code-fold: true

cor.test(alzheimer_healthy$age, alzheimer_healthy$hippo)

#alternatively, we can use linear regression
summary(lm(alzheimer_healthy$hippo ~ alzheimer_healthy$age))
```


# Example 7: one-way ANOVA

## Example 7: Visualizaiton

## Example 7: aov
```{r}
#| code-fold: true


summary(aov(alzheimer_data$lhippo ~ alzheimer_data$diagnosis))

#equivalently
summary(aov(lhippo ~ diagnosis, data=alzheimer_data))
```

