---
title: "Project - Normal Body Temperature"
author: "Zhaoxia Yu, Department of Statistics, University of California, Irvine"
date: "`r Sys.Date()`"
format: 
  revealjs:
    theme: "sky"
    slideNumber: true
    transition: "fade"
    progress: true
    controls: true
    code-fold: true
    echo: true
    fig-align: center
---

## Load Libraries
```{r}
library(ggplot2)
library(tidyverse)
```

## Background
- The original study was published in 2020:

- Protsiv, M., Ley, C., Lankester, J., Hastie, T. and Parsonnet, J., 2020. Decreasing human body temperature in the United States since the industrial revolution. Elife, 9, p.e49555.

## Background
- The article studied whether and how normal body temperature changed in the past 160 years. It examined three studies:
  - Union Army Veterans of the Civil War (UAVCW) Data: measurement years 1860–1940
  - National Health and Nutrition Examination Survey I: 1971–1975
  - Stanford Translational Research Integrated Database Environment: 2007–2017

## Your Project 

- In your project, please focus on the UAVCW data. 

- Based on the UAVCW data, please try to understand what factors were associated with normal body temperature between 1860 and 1940. 

- Be sure to interpret your results in a scientific context. 

- Conducting research on relevant literature is always a good idea.

- Conclude your presentation with a discussion to (1) highlight your main findings and (2) state any limitations of your study. 

## Read Data
```{r}
uavcw=read.csv('data4_temperature/veterans_processed.csv')
dim(uavcw)
par(mfrow=c(1,2))
plot(unlist(lapply(split(uavcw$temp_C, uavcw$examyear), mean)))
plot(temp_C ~ examyear, data=uavcw)

#do some exploratory analysis
#decide whether you would like to focus on a specific subset of data
```

