---
title: "Day 6 - Logistic Regression"
author: "Zhaoxia Yu, Department of Statistics, University of California, Irvine"
date: "`r Sys.Date()`"
format: 
  revealjs:
    scrollable: true
    theme: "sky"
    slideNumber: true
    transition: "fade"
    progress: true
    controls: true
    code-fold: true
    echo: true
    R.options:
      fig-align: center
---

## Load Libraries

```{r}
library(ggplot2)
library(tidyverse)
```

## Read Data and Create a Binary Variable

```{r}
#read, select variables
#remove impaired, and define new variables
alzheimer_data <- read.csv('data/alzheimer_data.csv') %>% 
  select(id, diagnosis, age, educ, female, height, weight, lhippo, rhippo) %>%
  filter(diagnosis!=1) %>%
  mutate(alzh=(diagnosis>0)*1, female = as.factor(female), hippo=lhippo+rhippo) 

table(alzheimer_data$alzh)
ggplot(data=alzheimer_data, aes(x=alzh)) +
  geom_bar()
```

# Motivation

## alzheimer vs hippocampal

```{r}
#| fig-align: center
ggplot(data=alzheimer_data, aes(x=hippo, y=alzh)) +
  geom_point() +
  geom_smooth(method='lm', se=FALSE) +
  labs(title="alzh vs hippo")
```

-   Any comments?

## Predict Binary Responses (wrong way)

-   Now let us do predictions using linear regression

```{r}
#| fig-align: center
alzh.lm=lm(alzh ~ hippo, data=alzheimer_data)
#add lm.pred variable
alzheimer_data$lm.pred=predict(alzh.lm)
#add red flags
alzheimer_data$lm.pred.red = alzheimer_data$lm.pred
alzheimer_data$lm.pred.red[alzheimer_data$lm.pred<1 &
                             alzheimer_data$lm.pred>0]=NA
ggplot(alzheimer_data) +
  geom_point(aes(x=alzh, y=lm.pred)) +
  geom_point(aes(x=alzh, y=lm.pred.red), col="red") +
  geom_hline(yintercept = c(0,1), col='red') +
  labs(x="true alzh", y="predicted by lm")
```

-   Is `alzh.lm` a good model?

## Binary Response Variables

-   Recall that `alzh` is a binary variable.

-   We tries linear regression: `alzh` $\sim$ `hippo`

    -   It does not fit well: the difference between fitted and observed is large for most data points.
    -   The predicted/ values for the binary response variable `alzh` can be greater than 1 or less than 0.

## How to Model Binary Responses?

-   Linear regression is not good choice for a binary response variable.

-   When the response variable is binary, such as `alzh`, how should we model it using one or multiple covarites?

-   Recall that a Bernoulli distribution can be used to characterize the behavior of a binary random variable.

# Bernoulli Distribution and Log-Odds

## How to Model Binary Responses?

-   Let $Y_i$ denote the response variable of the $i$th subject. When there is not covarites $x's$, we use Bernoulli distribution $$Y_i \sim Bernoulli(\pi)$$
    -   $P(Y_i=1)=\pi$.
    -   $P(Y_i=0)=1-\pi$.
    -   Here we use $\pi$ instead of $p$ to denote the success probability for better clarity.

## How to Model Binary Responses?

-   When there is one explanatory variable, denoted by $x_i$ for subject $i$, it is reasonable to assume that the `success` probability is a **function** of $x_i$, i.e., $$Y_i \sim Bernoulli (\pi(x_i))$$

-   How do we define $\pi(x_i)$? Does the following linear function work? $$\pi(x_i)=\beta_0 + \beta_1 x_i$$

## The Successful Probability $\pi(x_i)$

-   $\pi(x_i)$ is the probability for $Y_i$ being 1, i.e.,

$$\pi = E[Y_i].$$

-   One issue with $\pi( x_i)=\beta_0 + \beta_1 x_i$ is that the range can be any real value.

-   Because $\pi(x_i)$ is a probability, it is desirable to make sure that it is $\in [0,1]$.

-   How?

## Odds

-   In statistics, odds quantifies of the relative probabilities of an event and its complement

::: {style="font-size: 50%"}
$$odds=\frac{P(A)}{P(A^c)}=\frac{P(A)}{1-P(A)}.$$
:::

-   For a binary variable $Y$, the odds is

::: {style="font-size: 50%"}
$$odds = \frac{P(Y=1)}{P(Y=0)}$$
:::

-   The range of odds is $[0,\infty)$. Can you find transformation such as that the range is $(-\infty, \infty)$?

## Log-Odds and Logit Transformation

::: {style="font-size: 75%"}
-   $log \left( \frac{\pi_i}{1-\pi_i}\right)$ is called the **logit** function, i.e.,

$$logit(\pi_i)=log \left( \frac{\pi_i}{1-\pi_i}\right)= log \left(\frac{E[Y_i=1]}{E[Y_i=0]} \right).$$
:::

```{r}
#| fig-align: center
p=seq(0.01, 0.99, 0.01)
logit.p=log(p/(1-p))
plot(p, logit.p, type="l", main="the logit function")
```

# Logistic Regression
## Logistic Regression

-   Now we have found a reasonable way to connect $\pi$ to $x_i$

::: {style="font-size: 75%"}
$$log \left( \frac{\pi_i}{1-\pi_i}\right) = log \left(\frac{E[Y_i=1]}{E[Y_i=0]} \right)= \beta_0 + \beta_1 x_i$$
:::

-   In addition, we also have the mechanism to model uncertainty: $Y_i \sim Bernoulli(\pi_i(x_i))$.

-   This is known as **logistic regression**.

- Note: $\pi=\frac{e^{\beta_0 + \beta_1 x_i}}{1+e^{\beta_0 + \beta_1 x_i} }$.


## Compute the Estimates

-   For a linear regression model, we have closed-form solutions.

-   This is not true for a logistic regression.

-   Estimation of is typically conducted by maximizing the corresponding likelihood function.

-   $\hat\beta_0$ and $\hat\beta_1$ have to be obtained using numerical algorithms. How?

::: {style="font-size: 85%; color: lightgray"}
-   Iteratively re-weighted least squares (IRLS): the default method used by R
-   The Newton-Raphson algorithm
:::

## Logistic Regression in R

```{r}
alzh.logistic=glm(alzh ~ hippo, family="binomial", data=alzheimer_data)
summary(alzh.logistic) %>% coefficients()
```

- Note, `glm` stands for generalized linear models, which include linear regression and logistic regression as special situations.

-   $\hat\beta_0=6.093409, \hat\beta_1=-1.184699$.

-   How should we interpret these values?

# Interpretation
## Interpret Results

::: {style="font-size: 85%"}
-   Consider $x_i=x_0$, the log-odds of AD is

$$logit(\hat\pi_i) = \hat\beta_0 + \hat\beta_1 x_0$$

-   If we increase the explanatory by one unit, the log-odds becomes

$$logit(\tilde \pi_i) = \hat\beta_0 + \hat\beta_1 x_0 + \hat\beta_1$$

-   Therefore, $logit(\tilde \pi_i)-logit(\hat \pi_i)=\hat\beta_1=-1.184699$.
:::

## Interpret Results

-   We have obtained the change in log-odds associated with one-unit increase in $x$.

-   How can we convert it to changes in odds, which is more well understandable?

-   By the definition of `logit`, we have

::: {style="font-size: 85%"}
$$logit(\tilde\pi_i) - logit(\hat\pi_i)=log\frac{\tilde \pi_i}{1-\tilde \pi_i} -log \frac{\hat \pi_i}{1-\hat \pi_i}=log \frac{\frac{\tilde \pi_i}{1-\tilde \pi_i}}{\frac{\hat \pi_i}{1-\hat \pi_i}}=-1.184699$$
:::

## Interpret Results

-   Taking the exponent on both sides of the equation,

$$\frac{\frac{\tilde \pi_i}{1-\tilde \pi_i}}{\frac{\hat \pi_i}{1-\hat \pi_i}}=exp(-1.184699)=0.3058$$

-   The new odds is 30.58% of the odds.
-   In other words, if the hippocampal value increases by 1cc, the odds of AD decreased by 1-30.58%=69.42%.

## Confidence intervals
- The `confint` function in R can be used to find confidence intervals for log-odds. The default uses the estimated standard error. 
- We need to convert the results to odds
```{r}
confint.default(alzh.logistic)[2,]
exp(confint.default(alzh.logistic)[2,])
```

## Confidence intervals
```{r}
1-exp(confint.default(alzh.logistic))[2,]
```

- Recall that the estimated decrease of AD associated with one-unit increase of hippocampal volume is 69.42%. 

- A 95% confidence interval is [64.98%, 73.29%]. 

## P-value
```{r}
summary(alzh.logistic) %>% coefficients()
```

- The p-value is very small, about $9.0\times 10^{-66}$
  - indicating reject the null hypothesis $H_0: \beta_1=0$
  - suggesting that AD is significantly associated with hippocampal volume. 

## Predicted Value vs Observed Values
- Note, when using ``predict``, it is necessary to specify the correct type of prediction.
  - The default is on the scale of the linear predictors (log-odds).
  - The `response` type gives predicted probability. 

## Predicted Value vs Observed Values

```{r}
#| fig-align: center
plot(alzheimer_data$alzh, predict(alzh.logistic, type="response"),
     xlab="observed alzh", ylab="predicted alzh")
```

## $X$ vs Observed Values

```{r}
#| fig-align: center
plot(alzheimer_data$hippo, predict(alzh.logistic, type="response"),
     xlab="hippo volume", ylab="predicted alzh")
```

# With Multiple $X$'s
- Similar to linear regression, we can include multiple explanatory variables. 

- Connect $\pi_i$(i.e., $E[Y_i]$) to a linear function of the $\vec x_i$:      $$log\frac{\pi_i}{1-\pi_i} = \beta_0 + x_{i1} \times \beta_1+ … + x_{ip} \times \beta_p$$

- And $Y_i \sim Bernoulli (\pi)$. 

## Example
```{r}
alzh.glm.multi = glm(alzh ~ age + female + educ, family=binomial, data=alzheimer_data)
summary(alzh.glm.multi)$coefficients
```

## Example: Interpretation

::: {style="font-size: 90%"}
-   Consider the age variable. The estimated coefficient is 0.04015417.
    What information does it provide?
-   The estimated log-odds AD for subject $i$ is (or add a constant
    \textcolor{red}{determined by study design}, see the slides about retrospective studies)
    $$logit(\hat\pi_i) = \hat\beta_0 + \hat\beta_{age} age_i + \hat\beta_2 female_i + \hat\beta_3 educ_i$$
-   Let $\tilde \pi_i$ denote estimated log-odds after one year
    $$logit(\tilde\pi_i) = \hat\beta_0 + \hat\beta_{age} (age_i+1) + \hat\beta_2 female_i + \hat\beta_3 educ_i$$
:::

## Interpretation

-   The estimated change in log-odds
    $$logit(\tilde\pi_i) - logit(\hat\pi_i)=log\frac{\tilde \pi_i}{1-\tilde \pi_i} -log \frac{\hat \pi_i}{1-\hat \pi_i}=0.018138$$
-   Take exponential of both sides, we have
    $$\frac{\frac{\tilde \pi_i}{1-\tilde \pi_i}}{\frac{\hat \pi_i}{1-\hat \pi_i}} = exp(0.04015417)$$

## Interpretate Coefficients

-   The odds of AD in one year later is
    $exp(0.04015417)=$\textcolor{red}{1.040971 times} of the current odds.

-   When holding gender and education fixed, the estimated increase in odds of AD in a year is
    $e^{0.04015417}-1=4.097\%$

- A 95% confidence interval is [3.08%, 5.12%].
```{r}
confint.default(alzh.glm.multi)[2,]
exp(confint.default(alzh.glm.multi)[2,])-1
```
## Interpretate Coefficients
-   What if we are interested in the increase in odds of AD in ten years (everything else is fixed)?
-   The estimated increase in odds of AD in 10 years is
    $$e^{10*0.04015417}-1=49.4\%$$
-   A $95\%$ C.I. for 10-year increase in odds: [35.5%, 64.8%]. 

\tiny

```{r}
10*confint.default(alzh.glm.multi)[2,]
exp(10*confint.default(alzh.glm.multi)[2,])-1
```

## Significance and P-value
- Lastly, let's take a look at significance.
- All the three explanatory variables are significant.
```{r}
summary(alzh.glm.multi)$coefficients[-1,]
```
